NeuronUnit: Collaboratively Testing the Validity of Neuroscientific Models

### 2013-11-30 NeurounitFrontiers/Paper.tex

Test-driven methodologies have started to see success in neuroscience as well, in the form of \emph{modeling competitions}. 
During these competitions, competitors develop and parameterize models based on publicly-available training data and submit them to a central server. There, submitted model are provided hidden testing data which they must use to produce predictions. These predictions are validated using publicly available criteria to produce summaries of the relative merits of different models, just as a test suite summarizes the state of a software project. 
Such competitions continue to drive important advances and improve scientists' understanding of their fields. 
For example, the quantitative single neuron modeling competition (QSNMC) \citep{jolivet_quantitative_2008} investigates the complexity-accuracy tradeoff among reduced models of excitable membranes; 
the ``Hopfield'' challenge \citep{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and 
examples from other areas of biomedical research abound (e.g. \url{http://www.the-dream-project.org}).
% \begin{figure}%{r}{0.6\textwidth}
% %\vspace{-38px}
% \includegraphics[scale=0.75]{sciunit_overview.pdf}
% \caption{Proposal overview. 
% \small{i) The \textit{SciUnit} framework can generate discipline-specific libraries for tests generation and model interfacing.  
% \textit{NeuronUnit} is proposed and described here.  Experimental data guides test development; model generation informs and is informed by these \textit{SciUnit} libraries.  
% Test suites for OSB and QSNMC (described in the text) are being built. 
% %\vspace{-10px}
% \label{fig:sciunit_overview}
% \end{figure}
% \leavevmode
% \vspace{-0px}
\begin{figure}
\vspace{-45px}
\centering
\includegraphics[scale=0.6]{diagram1.pdf}
\vspace{-65px}
\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \citep{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \citep{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).}  
\label{fig:sciunit_overview}\vspace{-10px}
\end{figure}
\leavevmode

Each of these examples has leveraged \emph{ad hoc} infrastructure to support model validation. 
While the specific criteria used to evaluate models can vary widely between modeling domains, the underlying methodology is common and could be implemented once. 
Recognizing this, and inspired by unit testing practices, we have developed a discipline-agnostic framework for developing \emph{model validation test suites} called \textit{SciUnit} (available from \url{http://sciunit.scidash.org}) \todo{cite sciunit paper again here?}. 
In this paper, we will begin by detailing \textit{SciUnit}, focusing on examples from single-neuron physiology (Sec. \ref{sec:sciunit}).

\textit{SciUnit} contains validation logic common across scientific disciplines, but a particular discipline, such as neurophysiology, might have more specialized logic associated with it that can be shared amongst its sub-disciplines (e.g. hippocampal neurophysiology). We anticipate a collaborative workflow where this common logic is developed in common repositories on a social coding service, here \textit{Github}. For neurophysiology, we have developed such a repository, called \textit{NeuronUnit}. This repository contains common testing logic as well as bridges to existing informatics infrastructure to make it easy to import existing data and models. In particular, we will show how models described using \text{NeuroML} and provided freely by the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}) can be tested in a fully automated fashion using published data curated by the \textit{NeuroElectro Project} (NeuroElectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library (\url{http://neuralensemble.org/NeuroTools}) to extract features from model outputs (Sec. \ref{sec:neuronunit}). Particular sub-disciplines, like hippocampal neurophysiologists, can use this common logic to select relevant models and tests, forming particular test suites in common repositories. The end result of this workflow is a table like the one shown in Figure \ref{fig:sciunit_overview}, where submitted models can be compared based on the scores they achieve on different selected tests. These tables serve as a summary of the status of a modeling community, just as the results from a suite of unit tests serve as a summary of the status of a software project, and help scientists answer the questions mentioned above more easily and accurately.

@@

%Simple executable \emph{validation tests} that compute agreement between a model prediction and an experimental observation.  
We begin by building an example \textit{SciUnit} test suite that could be used in neurophysiology. 
Suppose we have collected data from an experiment where current stimuli (measured in pA) was delivered to neurons in some brain region, while the somatic membrane potential of each stimulated cell (in mV) was recorded and stored.  
A model claiming to capture this cell type's membrane potential dynamics must be able to accurately predict a variety of features observed in these data.

>>>>

Test-driven methodologies have started to see success in neuroscience as well, in the form of \emph{modeling competitions}. 
During these competitions, competitors develop and parameterize models based on publicly-available training data and submit them to a central server. There, submitted models are provided hidden testing data which they must use to produce predictions. These predictions are validated using publicly available criteria to produce summaries of the relative merits of different models, just as a test suite summarizes the state of a software project. 
Such competitions continue to drive important advances and improve scientists' understanding of their fields. 
For example, the quantitative single neuron modeling competition (QSNMC) \citep{jolivet_quantitative_2008} investigated the complexity-accuracy tradeoff among reduced models of excitable membranes; 
the ``Hopfield'' challenge \citep{hopfield_what_2000} tested techniques for generating neuronal network form given its function; 
the Neural Prediction Challenge sought the best stimulus reconstructions, given neuronal activity (\url{http://neuralprediction.berkeley.edu}); the Diadem challenge is advancing the art of neurite reconstruction (\url{http://www.diademchallenge.org}); and 
examples from other areas of biomedical research abound (e.g. \url{http://www.the-dream-project.org}).
% \begin{figure}%{r}{0.6\textwidth}
% %\vspace{-38px}
% \includegraphics[scale=0.75]{sciunit_overview.pdf}
% \caption{Proposal overview. 
% \small{i) The \textit{SciUnit} framework can generate discipline-specific libraries for tests generation and model interfacing.  
% \textit{NeuronUnit} is proposed and described here.  Experimental data guides test development; model generation informs and is informed by these \textit{SciUnit} libraries.  
% Test suites for OSB and QSNMC (described in the text) are being built. 
% %\vspace{-10px}
% \label{fig:sciunit_overview}
% \end{figure}
% \leavevmode
% \vspace{-0px}
\begin{figure}
\vspace{-45px}
\centering
\includegraphics[scale=0.6]{diagram1.pdf}
\vspace{-65px}
\caption{NeuronUnit overview. NeuronUnit is set of testing tools built upon the discipline-agnostic SciUnit framework. 
NeuronUnit can in principle test arbitrary neurophysiology models using arbitrary data but we provide here an example using models described in NeuroML as part of the \textit{Open Source Brain Project} (OSB, \citep{gleeson_open_2012}, \url{http://www.opensourcebrain.org}), and single neuron electrophysiology data available as part of the NeuroElectro Project (Neuroelectro, \citep{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}).}  
\label{fig:sciunit_overview}\vspace{-10px}
\end{figure}
\leavevmode

Each of these examples has leveraged \emph{ad hoc} infrastructure to support model validation. 
While the specific criteria used to evaluate models can vary widely between modeling domains, the underlying methodology is common and could be implemented once. 
Recognizing this, and inspired by unit testing practices, we have developed a discipline-agnostic framework for developing \emph{model validation test suites} called \textit{SciUnit} \citep{omar_sciunit_2013} (available from \url{http://sciunit.scidash.org}) \todo{cite sciunit paper again here?}. 
In this paper, we will begin by detailing \textit{SciUnit}, focusing on examples from single-neuron physiology (Sec. \ref{sec:sciunit}).

\textit{SciUnit} contains validation logic common across scientific disciplines.
But a particular discipline, such as neurophysiology, might have more specialized logic associated with it that can be shared amongst its sub-disciplines (e.g. hippocampal neurophysiology). 
We anticipate a collaborative workflow where this common logic is developed in common repositories on a social coding service, here \textit{Github}. 
For neurophysiology, we have developed such a repository, called \textit{NeuronUnit}. 
This repository contains common testing logic as well as bridges to existing informatics infrastructure to make it easy to import existing data and models. 
In particular, we will show how models described using \text{NeuroML} and provided freely by the \textit{Open Source Brain Project} (OSB, \cite{gleeson_open_2012}, \url{http://www.opensourcebrain.org}) can be tested in a fully automated fashion using published data curated by the \textit{NeuroElectro Project} (NeuroElectro, \cite{tripathy_neuroelectro:_2012}, \url{http://neuroelectro.org}), leveraging facilities from the \textit{NeuroTools} library (\url{http://neuralensemble.org/NeuroTools}) to extract features from model outputs (Sec. \ref{sec:neuronunit}). 
Scientists in particular sub-disciplines, like hippocampal neurophysiologists, can use this common logic to select relevant models and tests, forming focused test suites in common repositories. 
The end result of this workflow is a table like the one shown in Figure \ref{fig:sciunit_overview}, where submitted models can be compared based on the scores they achieve on different selected tests. 
These tables serve as a summary of the status of a modeling community, just as the results from a suite of unit tests serve as a summary of the status of a software project, and help scientists answer the questions mentioned in section \ref{sec:introduction} more easily and accurately.

@@

%Simple executable \emph{validation tests} that compute agreement between a model prediction and an experimental observation.  
We begin by building an example \textit{SciUnit} test suite that could be used in neurophysiology. 
Suppose we have collected data from an experiment where current stimuli (measured in pA) are delivered to neurons of a particular subtype, while the somatic membrane potential of each stimulated cell (in mV) is recorded and stored.  
A model claiming to capture this neuron type's membrane potential dynamics must be able to accurately predict a variety of features observed in these data.
